---
title: "BDA - Assignment X"
author: "Christian Segercrantz"
date: "9/15/2021"
output: 
  pdf_document: 
    toc: yes
    toc_depth: 1
urlcolor: blue
---

### 1. Basic probability theory notation and terms

* probability: The process used to predict what will happen (in the future).

* probability mass: The probability that a discrete random variable will have exactly some value.

* probability density: The relative (to other points) likelihood that a continuous random variable would take a given value.

* probability mass function (pmf): A function of the probability mass which gives us the probability that the random variable takes on a certain discrete value.

* probability density function (pdf): A function of the probability density which gives us the probability of the random variable falling in some range.

* probability distribution: The mathematical function which takes in the sample space of the variable and outputs the probabilities of events.

* discrete probability distribution: A probability distribution of a random variable that only take on discrete values.

* continuous probability distribution: A probability distribution of a random variable of real values.

* cumulative distribution function (cdf): The cumulative distribution function gives the probability that $P(X\leq x)$, i.e. the probability that the random variable takes on at least the value x. 

* likelihood: The probability of seeing a certain outcome of a underlying defined model.  



### 2. Basic computer skills

#### a

```{r}
rm(list = ls())
mu = 0.2
sigma2=0.01
alpha=mu*((mu*(1-mu)/sigma2)-1)
beta=alpha*(1-mu)/mu
x = seq(from=0,to=1, by=0.01)
betapdf = dbeta(x,alpha,beta)
plot(x, betapdf, type='l', main='Beta-distribution PDF')
```

#### b

```{r}
n = 1000
sample = rbeta(n,alpha,beta)
hist(sample)
```

We can see that the density function has a very similar look to the histogram.

#### c

```{r}
mean = mean(sample)
variance = var(sample)
```

We can see that the mean and the true mean are roughly very close to each other. The true variance \sigma\^2 and the variance are also very close to each.

#### d

```{r}
CI95=quantile(sample,probs=c(0.025,0.975))
plot(x,betapdf, type='l', main='Beta-distribution PDF')
abline(v = CI95[[1]], lty = 2)
abline(v = CI95[[2]], lty = 2)
```

### 3. Bayes’ theorem

```{r}
rm(list = ls())
P_test_pos_con_cancer_pos = 0.98
P_test_neg_con_cancer_neg = 0.96
P_cancer_pos = 1/1000
P_cancer_neg = 1 - P_cancer_pos
P_test_neg_con_cancer_pos = 1 - P_test_pos_con_cancer_pos
P_test_pos_con_cancer_neg = 1 - P_test_neg_con_cancer_neg
P_test_pos_and_cancer_pos = P_test_pos_con_cancer_pos * P_cancer_pos
P_test_neg = P_test_neg_con_cancer_neg * P_cancer_neg + P_test_neg_con_cancer_pos * P_cancer_pos
P_test_pos = 1-P_test_neg
P_cancer_pos_con_test_neg = (P_test_neg_con_cancer_pos * P_cancer_pos) /  P_test_neg
P_cancer_pos_con_test_pos = P_test_pos_con_cancer_pos * P_cancer_pos/ P_test_pos
cat(paste0("P(cancer=positive|test=neg)=",round(P_cancer_pos_con_test_neg*100,4),"%"),sep="")
cat(paste0("P(cancer=positive|test=positive)",round(P_cancer_pos_con_test_pos*100,4),"%"),sep="")
```

We compute the following numbers: P(cancer=positive\|test=neg) = 0.002% P(cancer=positive\|test=positive) = 2.3% This means that so called false negatives are rare, but true positives are also very low. This means that very few of those who test positive actually have cancer only 2,3% meaning that it will send approximately 50 times too many people to further tests and thus does not do what it's planned to do. This happens since the percentage of people actually having cancer is very small.

### 4. Bayes’ theorem

#### a

```{r}
rm(list = ls())
p_red = function(boxes){
  P_A = 0.4
  P_B = 0.1
  P_C = 1 - P_A - P_B
  P_red_con_A = boxes[1,1]/rowSums(boxes)[1]
  P_red_con_B = boxes[2,1]/rowSums(boxes)[2]
  P_red_con_C = boxes[3,1]/rowSums(boxes)[3]
  res = P_red_con_A*P_A + P_B*P_red_con_B + P_red_con_C * P_C
  return(res)
}

boxes <- matrix(c(2,4,1,5,1,3), ncol = 2,dimnames = list(c("A", "B", "C"), c("red", "white")))
p_red(boxes = boxes)


```

There is a approximately 28.4% of picking a red ball.

#### b

```{r}
p_box = function(boxes){
  P_A = 0.4
  P_B = 0.1
  P_C = 1 - P_A - P_B
  P_red_res = p_red(boxes = boxes)
  P_red_con_A = boxes[1,1]/rowSums(boxes)[1]
  P_red_con_B = boxes[2,1]/rowSums(boxes)[2]
  P_red_con_C = boxes[3,1]/rowSums(boxes)[3]
  P_A_con_red = P_red_con_A * P_A / P_red_res
  P_B_con_red = P_red_con_B * P_B / P_red_res
  P_C_con_red = P_red_con_C * P_C / P_red_res
  res = c(P_A_con_red,P_B_con_red,P_C_con_red)
  return(res)
}
p_box(boxes = boxes)
```

### 5. Bayes’ theorem

#### a

```{r}
p_identical_twin = function(fraternal_prob, identical_prob){
  P_boy = 1/2
  P_ident_and_twin_boy = P_boy*identical_prob
  P_frat_and_twin_boy = P_boy*P_boy*fraternal_prob
  P_ident_con_twin_boy = P_ident_and_twin_boy/(P_ident_and_twin_boy+P_frat_and_twin_boy)
  return(P_ident_con_twin_boy)
}
p_identical_twin(fraternal_prob = 1/150, identical_prob = 1/400)

```

```{r}
library(markmyassignment)
assignment_path <-  paste("https://github.com/avehtari/BDA_course_Aalto/","blob/master/assignments/tests/assignment1.yml", sep="")
set_assignment(assignment_path)
#show_tasks()
mark_my_assignment()

```
