---
title: "BDA - Assignment 4"
date: "10/10/2021"
output: 
  pdf_document: 
    toc: yes
    toc_depth: 2
urlcolor: blue
---

```{r setup, include=FALSE, echo=FALSE}
rm(list = ls())
setwd("~/GitHub/Beysian-Data-Analysis-course-F2021/Assignment 4")
library(aaltobda)
data("bioassay_posterior")
data("bioassay")
```

# a)

```{r}
mean = c(0,10)
corr = 0.6
a_std = 2
b_std = 10

cov = matrix( c(a_std^2, a_std*b_std*corr, a_std*b_std*corr, b_std^2),nrow = 2)
mean
cov
```

The mean vector and the covariance matrix can be seen printed by the last cell.

# b)

```{r}
alpha_mean = mean(bioassay_posterior$alpha)
beta_mean = mean(bioassay_posterior$beta)

alpha_interval = quantile(probs = c(0.05, 0.95), bioassay_posterior$alpha)
beta_interval = quantile(probs = c(0.05, 0.95), bioassay_posterior$beta)

alpha_mean_MSCE = sqrt(var(bioassay_posterior$alpha)/length(bioassay_posterior$alpha))
beta_mean_MSCE = sqrt(var(bioassay_posterior$beta)/length(bioassay_posterior$beta))

alpha_interval_MSCE = mcse_quantile(bioassay_posterior$alpha, 0.9)
beta_interval_MSCE = mcse_quantile(bioassay_posterior$beta, 0.9)
```

The mean alpha is `r round(alpha_mean, digits = 1)`.0 with MSCE `r round(alpha_mean_MSCE, digits=3)` and beta `r round(beta_mean, digits = 2)` with MSCE `r round(beta_mean_MSCE, digits=3)`.

The qunatile interval for alpha is [`r round(alpha_interval, digits = 1)`] with a MSCE of `r round(alpha_interval_MSCE, digits=3)` and for beta [`r round(beta_interval, digits = 0)`] with a MSCE of `r round(beta_interval_MSCE, digits=3)`.

<!-- Explain in words something -->

# c)

```{r}
log_importance_weights = function(alpha, beta){
  bioassaylp(alpha, beta, bioassay$x, bioassay$y, bioassay$n)
}
```

```{r include=FALSE, echo=FALSE}
#test case
alpha <- c(1.896, -3.6, 0.374, 0.964, -3.123, -1.581)
beta <- c(24.76, 20.04, 6.15, 18.65, 8.16, 17.4)
round(log_importance_weights(alpha, beta),2)
```

```{r}
liw = log_importance_weights(bioassay_posterior$alpha, bioassay_posterior$beta)
```

The some of the log importance weights are: `r head(liw)`.

<!-- Explain in words -->

# d

```{r}
normalized_importance_weights = function(alpha, beta){
  liw = log_importance_weights(alpha,beta)
  exp_liw = exp(liw)
  norm_exp_liw = exp_liw/sum(exp_liw)
  return(norm_exp_liw)
}
```

```{r include=FALSE, echo=FALSE}
round(normalized_importance_weights(alpha = alpha, beta = beta),3)
```

```{r}
niw = normalized_importance_weights(bioassay_posterior$alpha,bioassay_posterior$beta)
```
The some of the nromalized importance weights are: `r head(niw)`.


# e)

```{r}
n=4000
draws = rmvnorm(n, mean, cov)
normal_niw = normalized_importance_weights(draws[,1], draws[,2])
hist(normal_niw)
```

```{r}
  S_eff = function(alpha, beta){
  niw = normalized_importance_weights(alpha,beta)
  1/sum(niw^2)
}
```

```{r include=FALSE, echo=FALSE}
round(S_eff(alpha = alpha, beta = beta),3)
```
```{r}
isess = S_eff(alpha = draws[,1], beta = draws[,2])
```

The importance sampling effective sample size is `r round(isess,digits=3)`.

# g)

# h)

```{r}
posterior_mean = function(alpha, beta){
  seff = S_eff(alpha,beta)
  
  niw = normalized_importance_weights(alpha,beta)
  
  alpha_mean = mean(niw*alpha)/mean(niw)
  beta_mean = mean(niw*beta)/mean(niw)

  alpha_var = mean(niw*alpha^2)/mean(niw) - alpha_mean^2
  beta_var = mean(niw*beta^2)/mean(niw) - beta_mean^2
  
  alpha_mean_MSCE = sqrt(alpha_var/seff)
  beta_mean_MSCE = sqrt(beta_var/seff)
  res = list(alpha_mean = alpha_mean, beta_mean = beta_mean, alpha_mean_MSCE = alpha_mean_MSCE, beta_mean_MSCE = beta_mean_MSCE)
  return(res)
}
```

```{r echo=FALSE,include=FALSE}
posterior_mean(alpha = alpha, beta = beta)
```

```{r}
post_mean = posterior_mean(draws[,1],draws[,2])
```

The posterior mean for alpha is `r round(post_mean$alpha_mean, digits=1)`.0 with a MCSE of `r round(post_mean$alpha_mean_MSCE, digits=3)` and beta is `r round(post_mean$beta_mean, digits=0)` with MCSE of  `r round(post_mean$beta_mean_MSCE, digits=3)`. 

```{r echo=FALSE, include=FALSE}
library(markmyassignment)
assignment_path <-
paste("https://github.com/avehtari/BDA_course_Aalto/",
"blob/master/assignments/tests/assignment4.yml", sep="")
set_assignment(assignment_path)
# To check your code/functions, just run
mark_my_assignment()
```

