---
title: "BDA - Assignment 5"
date: "17/10/2021"
output: 
  pdf_document: 
    toc: yes
    toc_depth: 2
urlcolor: blue
---

```{r include=FALSE, echo=FALSE}
rm(list = ls())
setwd("~/GitHub/Beysian-Data-Analysis-course-F2021/Assignment 5")
library(aaltobda)
data("bioassay")
library("rstan")
```

# 1

We start by pre-calculating the mean $\mu$ and covariance metrics $\Sigma$

```{r}
corr = 0.6
a_std = 2
b_std = 10

mu = c(0,10)
sigma = matrix( c(a_std^2, a_std*b_std*corr, a_std*b_std*corr, b_std^2),nrow = 2)
```

We continue

```{r}
density_ratio = function(alpha_propose, alpha_previous, beta_propose, beta_previous, x, y, n){
  likli_propose <- bioassaylp(alpha_propose, beta_propose, x, y, n)
  likli_previous <- bioassaylp(alpha_previous, beta_previous, x, y, n)
  
  prior_prop <- dmvnorm(c(alpha_propose,beta_propose),mu, sigma, log = TRUE)
  prior_prev <- dmvnorm(c(alpha_previous,beta_previous),mu, sigma, log = TRUE)
  
  prop = prior_prop + likli_propose
  prev = prior_prev + likli_previous
  
  res = exp(prop - prev)
  return(res)
}

```

```{r, include=FALSE, echo=FALSE}
#Test1
density_ratio(alpha_propose = 1.89, alpha_previous = 0.374,
beta_propose = 24.76, beta_previous = 20.04,
x = bioassay$x, y = bioassay$y, n = bioassay$n)
```

```{r, include=FALSE, echo=FALSE}
#Test2
density_ratio(alpha_propose = 0.374, alpha_previous = 1.89,
beta_propose = 20.04, beta_previous = 24.76,
x = bioassay$x, y = bioassay$y, n = bioassay$n)
```

```{r}
Metropolis_bioassay = function(n, alpha_previous, beta_previous, warmup_procent = 0.5){
  alphas = c()
  betas = c()
  for( i in 1:n){
    alpha_propose = rnorm(1, alpha_previous, 1)
    beta_propose = rnorm(1, beta_previous, 5)
    
     r = density_ratio(alpha_propose = alpha_propose, alpha_previous = alpha_previous, beta_propose = beta_propose, beta_previous = beta_previous, x = bioassay$x, y = bioassay$y, n = bioassay$n)
     r = min(1,r)
     if(r >= runif(1)){
       alpha_previous = alpha_propose
       beta_previous = beta_propose
     }
    alphas[i] = alpha_previous
    betas[i] = beta_previous
  }
  alpha_final = na.omit(alphas[n*warmup_procent+1:n])
  beta_final = na.omit(betas[n*warmup_procent+1:n])
  return(cbind(alpha_final,beta_final))
}
```


```{r}
n=2000
warmup_procent = 0.5
thetas1 = Metropolis_bioassay(n, 0, 10, warmup_procent)
thetas2 = Metropolis_bioassay(n, 2, 15, warmup_procent)
thetas3 = Metropolis_bioassay(n, -2, 5, warmup_procent)
```

We will simulate 3 different chains using different initial starting points. The points are chosen as the priors means as well as one standard deviations in both positive and negative direction. This gives us the starting points $\alpha_{init} = [0,2,-2]$ and $\beta_{init} = [10,15,5]$. We draw $n=$ `r n` draws for each chain and use a warm-up length of half the chain, i.e. $n_{warmup}=$ `r n*warmup_procent`.

```{r}
plot(thetas1[,1],
     type='l',
     col='red',
     ylab='alpha')
lines(thetas2[,1],
     type='l',
     col='blue')
lines(thetas3[,1],
     type='l',
     col='black')
```

```{r}
plot(thetas1[,2],
     type='l',
     col='red',
     ylab='beta')
lines(thetas2[,2],
     type='l',
     col='blue')
lines(thetas3[,2],
     type='l',
     col='black')
```

```{r}
alphas = cbind(thetas1[,1],thetas2[,1],thetas3[,1])
alpha_Rhat = Rhat(alphas)
betas = cbind(thetas1[,2],thetas2[,2],thetas3[,2])
betas_Rhat = Rhat(betas)
```

The values are $\widehat R_\alpha=$ `r round(alpha_Rhat,digits=3)` and $\widehat R_\beta=$ `r round(alpha_Rhat,digits=3)`.

```{r}

plot(thetas1[,1], thetas1[,2],
     col='red',
     xlab='alpha',
     ylab="beta",
     main="Draws of the parameters")
points(thetas2[,1], thetas2[,2], col='red')
points(thetas3[,1], thetas3[,2], col='red')

```
