---
title: "BDA - Assignment 6"
date: "24/10/2021"
output: 
  pdf_document: 
    toc: no
urlcolor: blue
---

```{r include=FALSE, echo=FALSE}
rm(list = ls())
setwd("~/GitHub/Beysian-Data-Analysis-course-F2021/Assignment 6")
library(aaltobda)
library("rstan")
data("bioassay")
```

# 1

We start by pre-calculating the mean $\mu$ and covariance metrics $\Sigma$

```{r}
corr = 0.6
a_std = 2
b_std = 10

mu = c(0,10)
sigma = matrix( c(a_std^2, a_std*b_std*corr, a_std*b_std*corr, b_std^2),nrow = 2)
```

Here is the stan model used for the assignment

```{stan, output.var="stanmodel"}
data {
  int<lower=0> N;
  vector[N] x;
  int<lower=0> n[N];
  int<lower=0> y[N];
  vector[2] mu; 
  matrix[2,2] sigma;
}
parameters {
  vector[2] theta; 
}
model {
  theta ~ multi_normal(mu, sigma);
  for (k in 1:N) {
    y[k] ~ binomial_logit(n[k],theta[1] + theta[2]*x[k]);
  }
  
}
```

Let us create the data used by stan 

```{r, message=FALSE, warning=FALSE}
data <- list(
  N = length(bioassay$x), #Number of data points
  x = bioassay$x,         #Outcome
  n = bioassay$n,         #Total draws
  y = bioassay$y,         #Successes
  mu = mu,                #Mean vector
  sigma = sigma           #Covariance matrix
)
```

```{r}
fit1 = sampling(stanmodel,
  data = data,            # named list of data
  chains = 4,             # number of Markov chains
  warmup = 1000,          # number of warmup iterations per chain
  iter = 2000,            # total number of iterations per chain
  cores = 1,              # number of cores (could use one per chain)
  refresh = 0             # no progress shown
  )
```
We have chosen to use a chain length of 2000 with a warmup of 1000. We are simulating 4 chains.

Let us print the model
```{r}
print(fit1)
```

and here we can see how the parameters alpha and beta are distributed

```{r, message=FALSE, warning=FALSE}
plot(fit1, probs=c(.05,.5,.95))
```
We can now plot all chains to examine graphically if they converge.
```{r}
traceplot(fit1, inc_warmup = TRUE, nrow = 2)
```

# 2

We use the built in function for Rhat from Rstan

```{r}
diagnostics = monitor(fit1)
```

The $\widehat R_{\alpha}$ is `r round(diagnostics$Rhat[1],digits=2)`.00 and $\widehat R_{\beta}$ is `r round(diagnostics$Rhat[2],digits=2)`.00. We can from this conclude that since the $\widehat R$-values are $\widehat R < 1.05$ that the chains have converged well.

# 3
Let us scatter plot the parameters
```{r}
theta = extract(fit1)$theta
alpha = theta[,1]
beta = theta[,2]
plot(alpha,beta)
```
We can conclude that the plot is very similar to last weeks plot and the plot from the course book.

# 4. 

The following tech was used for the assignment:

1. OS: Windows 10 

2. Language: R

3. Interface used: Rstan

I had no installation or compilation error. I ran everything locally. I found the documentation quite lacking, I found no explanation for the `y ~ dist(para1,para2)` kind of syntax and when to use it and when to use normal `y = dist(y,para1,para2)` syntax. 
